Towards Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-26 manuscript/manuscript.tex
Finish edits

\title[Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022]{Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022}

@@

\textbf{Results:} GPT-4 had the best performance overall (0.61 PCCC), followed by GPT-3.5 (0.56 PCCC), and InSilicoVA/InterVA-5 (0.44 PCCC). GPT-4 had the best performance for adult and neonatal records (0.64 and 0.58 PCCC), with GPT-3.5 for child records (0.54 PCCC). All models' performances trended upwards from 1 month to 14 years ($~$0.1-0.75 PCCC) and downwards from 15-69 years ($~$0.7-0.35) of age. GPT-4, GPT-3.5, ad InSilicoVA had the highest performances for 17, 9, and 4 of all 30 CODs respectively. At the population level, all models had CSMF accuracies between 0.7-0.79.

@@

This study uses 6939 physician agreed records (two physicians with similar COD assignment on the same record) of 11,920 records collected in 2019 to 2022 from the Healthy Sierra Leone (HEAL-SL) study as described in Section \ref{methods-data}. Section \ref{methods-models} describes the four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, for COD assignment, and their inputs and outputs, while section \ref{methods-eval} details the performance evaluation of the four models relative to physicians overall, by ages, and by COD, using population and individual level metrics. See Appendix \ref{app-methods} for additional details on the methods used in this study.

@@

The performance of the four models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. As model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}, the CSMF accuracy and PCCC metrics were compared for each model overall, by age group (adult, child, neonatal), by CGHR-10 COD codes, and across age ranges. For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. See Appendix \ref{app-methods-eval} for more details on performance metrics and evaluation strategy for comparing each model.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Across age ranges, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}). PCCC trended upwards for 1 month to 14 years ($~$0.1-0.75), and downwards for ages 15 to 69 years ($~$0.7-0.35). The highest and lowest performances were observed for ages 10-29 years ($~$) and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

At the population level, overall performances for all models were similar to physicians (0.74-0.79 CSMF), indicating potential for adequately estimating COD distributions for large populations. Although all models did not perform well for all records at the individual level (0.44-0.61 PCCC), several models performed well for certain CODs (0-0.99 PCCC). For most CODs, GPT-3.5/GPT-4 performed better than InSilicoVA/InterVA-5 (top PCCC for 15 of 17, 7 of 8, and 4 of 5 adult, child, and neonatal CODs respectively), while InSilicoVA performed better for particular CODs (road and transport injuries, tuberculosis, pneumonia, and neonatal infections with 0.84, 0.71, 0.78, and 0.86 PCCC respectively). For CODs with high performance (e.g. GPT-3.5/4 with 0.91-0.99 PCCC for maternal conditions, chronic respiratory disease, diabetes melitus, and cancers, InSilicoVA with 0.84 and 0.86 PCCC for road and transport injuries, and neonatal infections), the results suggest that GPT-3.5/4 and InSilicoVA may assign CODs that are very similar to physicians. Thus, it may be beneficial to evaluate performance at the COD level, and apply a combination of models that perform well in comparison to physicians for each COD. For example, different models perform well for various leading CODs as seen in Table \ref{tab-leadcod} \cite{worldhealthorganizationWorldHealthStatistics2024,worldhealthorganizationCausespecificMortality2000}.

@@

Across age ranges, all models followed a similar upward trend from 6 months to 14 years of age, and a downward trend from 15-69 years with GPT models having higher performance than InSilicoVA/InterVA-5 models, while more mixed trends were observed from 0 days to 5 months (recall Figure \ref{fig-pccc-age}). For adult age ranges, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults with some improvements after the age of 59. For child and neonatal age ranges, the performance improves drastically as the age increases after 5 months, suggesting less difficulty in COD assignment when children and neonates are more developed. As the models did not perform particularly well ($\geq$ 0.8 PCCC) for any specific five-year age range, it is not recommended to apply specific models that target cases by age. However, the patterns of increases and decreases of performance in relation to age provide valuable insight for comparison to expected physician diagnosis patterns in well-studied medical literature and knowledge. For example, it may be expected that physicians are more uncertain in diagnosing diseases that are prevalent in neonatal patients \cite{rasmussenComplexityPhysiciansUnderstanding2019,faisonWhenUnknownUnknowable2023}, which are present in our findings from Figure \ref{fig-pccc-age}.

@@

The GPT models in this study had the temperature parameter set to 0 for more reproducible results. A short experiment in Appendix \ref{app-reprod} revealed that GPT-3.5 assigns the same COD for the same record only more than 60\% of the time, based on repeated runs on a sample of 100 records. Although more extensive testing is needed, this suggests that GPT models do not always assign the COD for the same case on multiple runs, which may pose issues in reproducibility and reliability. For example, GPT models may achieve correct COD assignments solely due to random chance, but are difficult to test with large numbers (e.g. 10,000) of reruns due to costs (e.g. costs increased 10 fold per record when rerun 10 times). In comparison, InterVA-5 and InSilicoVA are open source and free, and assign deterministic CODs with probabilities for each alternative COD, which offers more reproducible and reliable COD assignments despite lower performance overall. In addition, all models were trained on historical data up to particular points in time, which may not utilize the most up-to-date data available (e.g. latest online articles, social media, or books for GPT models).

@@

GPT-3.5 and GPT-4 models required large computing infrastructure to train and run, which was not possible to run on local computers, or setup due to costs. This poses issues with data privacy as sensitive data (e.g. identifying information) needs to be sent to company servers, which can be collected by companies (e.g. OpenAI) and misused \citep{taoOpeningPandoraBox2023}. For example, GPT models use prompts, which contain the narrative data, to assign CODs. The data in the promots may be unknowingly collected and misused by companies (e.g. companies) or their users (e.g. malicious prompts) \citep{khowajaChatGPTNeedsSPADE2024,wuUnveilingSecurityPrivacy2024}. In contrast, InterVA-5 and InSilicoVA can be run on local computers, which allows data to stay with the owner to protect data privacy, without reliance on external services.

@@

This section identifies limitations in this research in the context of GPT models. Section \ref{discuss-limits-icd10} identifies the omission of ICD-10 performance evaluations. Section \ref{discuss-limits-data} discusses data limitations, such as small samples, exclusion of disagreements, and exploration of incorrectly assigned cases. Section \ref{discuss-limits-model} mentions the need for parameter tuning and evaluation of consistency and multiple COD assignments.

\subsubsection{ICD-10 Performance Evaluation}\label{discuss-limits-icd10}

For the scope of this study, all models were evaluated for their performance in broad CGHR-10 COD categories as opposed to more specific ICD-10 codes. However, in practical cases, physicians assign more specific ICD-10 codes rather than broader COD categories. InterVA-5 and InSilicoVA assigned broader WHO VA codes, and were unable to assign ICD-10 codes, as the number of cases for specific ICD-10 codes may not have enough cases for training statistical models. GPT models were able to assign ICD-10 codes, but may possibly result in lower performance as even physicians do not agree completely on ICD-10 codes, and broader categories (CMEA-10 codes in Additional file 2) were used to assign equivalency or agreement.

\subsubsection{Small Samples, Disagreements, and Misclassifications}\label{discuss-limits-data}

Performance evaluations and analyses were omitted for CODs with less than 10 cases (e.g. congenital anomalies, suicide), in which models may perform well on, but have limited evidence to make conclusive insights on performance. In this study, only records where physicians agreed on the COD assignment were included in performance evaluations. However, the performance of GPT-4 only dropped slightly from 0.61 (agreed) to 0.53 (disagreed) PCCC while records almost doubled, which suggests that GPT models may potentially be able to assign agreeable CODs without reconciliation or adjudication for cases where physicians disagreed (Figure \ref{fig-perf-allvsagree}). In relation, GPT may be used to explore prompts that may possibly reconcile or adjudicate records in disagreement. Finally, an exploration of misclassifications was not conducted for this study, but may yield useful insights. For example, a non-comprehensive exploration was conducted in Appendix \ref{app-misclass} on misclassified GPT-4 records for neonatal infections, which found potential issues with the categorization of CGHR-10 codes, order of information in narratives, and guidelines of COD assignments. More records may also provide more conclusive evidence of model performance results as 6939 records may be inadequate evidence at the country level compared to more comprehensive studies (e.g. the million death study \citep{jhaProspectiveStudyOne2005}).

@@

GPT-3.5 and GPT-4 models used default parameters with the exception of setting the temperature to 0 for more consistent results. However, the temperature and other potential settings may be adjusted to possibly improve performance \citep{openaiOpenAIPlatformAPI2024}. In addition, GPT models may possibly produce inconsistent results even with the temperature set to 0 as discussed in Section \ref{discuss-disadv-reprod}. Thus, it is important to also test the reliability and consistency of GPT outputs to avoid coincidental results due to randomness \citep{johnsonAssessingAccuracyReliability2023,jangConsistencyAnalysisChatGPT2023,krishnaEvaluationReliabilityRepeatability2024}. InterVA-5 and InSilicoVA were able to provide multiple COD assignments with probabilities for each. GPT models can be prompted to produce more than one COD assignment, but was not explored in this study. This may be useful to evaluate the performance of suggested alternative COD assignments, which can help physicians reach agreement or reduce ill-defined assignments.

@@

Prompt engineering, the act of designing prompts to guide GPT models for better results \citep{wangPromptEngineeringHealthcare2024}, presents an important research opportunity that may improve performance of GPT models for COD assignment. As mentioned in Section \ref{discuss-limits-data} and exemplified in Appendix \ref{app-misclass}, an analysis of misclassified records may yield insights on adjusting prompts to assign more correct CODs in relation to physicians. In addition, subsequent prompts and examples can be used to add in correctional instructions and refine results, while additional information from the questionnaire and physician VA manuals may add better context \citep{meskoPromptEngineeringImportant2023}. Sensitivity analyses may be conducted to assess the effects on performance and consistency of results from modified prompts on a COD basis. GPT models may also be customized to specific domains or contexts, where objectives, behaviours, extra data, privacy, and evaluation tests can be adjusted to produce custom models that perform better in targeted domains or circumstances \citep{almasreDevelopmentEvaluationCustom2024}.

\subsubsection{Guided Household Surveys}\label{discuss-opp-gsurveys}

Finally, GPT models may also guide surveyors during VA interviews to probe households for better narrative information by suggesting better questions or questions that may have been missed by the surveyors. Lastly, as models can assign CODs on-demand, there is potential for models to provide immediate COD estimates during the data collection process to monitor data quality (e.g. comparing estimated to expected COD distributions for known areas as quality checks).

@@

Another research opportunity is in the integration of GPT, InterVA-5, and InSilicoVA models into VA systems to assist physicians in COD assignment. In dual-coded VA systems (described in Section \ref{methods-data}), two physicians are randomly assigned to each record and require second inspections of each other's assignment (reconciliation) and evaluation by a third more senior physician if their assignments do not agree. As mentioned in Section \ref{discuss-adv-hscav}, suggestion of alternative assignments from GPT and InSilicoVA models potentially reduces the disagreement between physicians, and ill-defined records, while allowing physicians to focus on more difficult records. Thus, model suggestions can be integrated into VA systems by presenting COD suggestions to physicians after their initial COD assignment, which allows them to consider alternative assignments and possibly revise their assignments based on the suggestions. At step 2 in Figure \ref{fig-discuss-phyassist}, GPT, InterVA-5, and InSilicoVA models can suggest COD assignments to consider, providing the option in step 2b to revise or proceed with their initial assignment. Our future work will be a first step in computer assisted verbal autopsy, assessing the effects of these model suggestions on improve VA data quality (e.g. increase in agreed records, reduction of ill-defined deaths). In preparation, we have integrated GPT-4, InterVA-5, and InSilicoVA model suggestions into our on-going HEAL-SL study after survey round 2 \citep{njalauniversityHealthySierraLeone2023} with goals of increasing physician agreement and reducing ill-defined COD assignments.

@@

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models had CSMF accuracies of 0.7-0.79. At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.45 PCCC). When evaluating performance by COD, 7 of 19, 4 of 10, and 1 of 7 adult (12-69 years), child (28 days to 11 years),
and neonatal (under 28 days) CODs respectively had models with $\geq$0.8 PCCC (see Table \ref{tab-perf-geq80}). Performance decreased from 0.72 to 0.59 PCCC as adults aged, and increased from 0.5 to 0.7 PCCC as children and neonates developed. Thus, GPT and InSilicoVA models were comparable to physicians for particular CODs (e.g. maternal conditions, injuries, diarrhoeal diseases, pneumonia), but not across age ranges. Advantages of GPT (and InSilicoVA for some CODs) models include being highly scalable and available, which allows the suggestion of multiple alternative COD assignments, and reduction of time spent on non-complex cases to assist physicians in COD assignment. In addition, GPT models provide flexible natural language input and output, capturing latent patterns (e.g. health-seeking and social issues) that potentially lead to their overall high performance compared to InterVA-5 and InSilicoVA. However, GPT models do not always assign the same COD for the same record on multiple runs, are trained on data from a specific time period, and require large computing infrastructure, leading to disadvantages in reliability of COD assignments, timeliness of up-to-date information, and data privacy issues. Limitations of this study were small sample sizes in relation to rarer CODs and other national level studies, need for more comprehensive exploration of disagreed and misclassified records, and lack of experimentation of model parameters, output consistency, and multiple COD assignments. Research opportunities include refining GPT models using prompt engineering and custom models, and future work towards computer assisted verbal autopsy, where GPT and other models are used to assist physician COD assignment by offering multiple alternative assignments, improving physician agreement on COD assignment and reducing ill-defined deaths. GPT-4, InterVA-5, and InSilicoVA has been integrated into future rounds of the HEAl-SL study to assist physicians with alternative COD assignments. Future work in evaluating the effectiveness of these models to reduce disagreements among physicians and ill-defined deaths will be a step forward in more accurate and efficient VA systems.

@@

\begin{figure}[H]
\centering
\includegraphics[width=.7\textwidth]{fig-methods.pdf}
\caption{Case study methods.}\label{fig-methods}
\end{figure}

>>>>

\title[Towards Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022]{Towards Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022}

@@

\textbf{Results:} GPT-4 had the best performance overall (0.61 PCCC), followed by GPT-3.5 (0.56 PCCC), and InSilicoVA/InterVA-5 (0.44 PCCC). GPT-4 had the best performance for adult and neonatal records (0.64 and 0.58 PCCC), with GPT-3.5 for child records (0.54 PCCC). All models' performances trended upwards from 1 month to 14 years ($\sim$0.1-0.75 PCCC) and downwards from 15-69 years ($\sim$0.7-0.35) of age. GPT-4, GPT-3.5, and InSilicoVA had the highest performances for 17, 9, and 4 of all 30 CODs respectively. At the population level, all models had CSMF accuracies between 0.74-0.79.

@@

This study details the methods used to compare the COD assignment from four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, to physicians as seen in Figure \ref{fig-methods-brief}. The initial VA data was filtered for physician agreed records as described in Section \ref{methods-data}. Section \ref{methods-models} describes the input and output of the four models for COD assignment, while section \ref{methods-eval} details the performance evaluation of the models relative to physicians using population and individual level metrics. See Appendix \ref{app-methods} for additional details on the methods used in this study.

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{fig-methods-brief.pdf}
\caption{Case study methods.}\label{fig-methods-brief}
\end{figure}

@@

The performance of the four models were evaluated with metrics at the population and individual level by comparing their CGHR-10 COD outputs for 6939 records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level (see Appendix \ref{app-methods-csmfa}), while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level (see Appendix \ref{app-methods-pccc}) \citep{murrayRobustMetricsAssessing2011}. Both CSMF accuracy and PCCC metrics are between 0 and 1 with 0 indicating low performance and 1 indicating perfect performance at the population and individual level respectively. As model performance can vary across ages and specific causes \citep{benaraEvaluationMethodsAssigning2023,murrayUsingVerbalAutopsy2014,setelValidityVerbalAutopsy2006}, the CSMF accuracy and PCCC metrics were compared for each model overall, by age group (adult, child, neonatal), by CGHR-10 COD codes, and across age ranges. For each of the adult and child age groups, metrics were calculated for five-year age ranges for records with ages at death of one-year or older and five-month age ranges for 28 days or older. For the neonatal age group, the age ranges of 0-6 days and 7-27 days were used. See Appendix \ref{app-methods-eval} for more details on performance metrics and evaluation strategy for comparing each model.

@@

Of all 6939 records, GPT-4 (0.61 PCCC) had the highest individual performance followed by GPT-3.5 (0.56 PCCC), InSilicoVA (0.44 PCCC), and InterVA-5 (0.44 PCCC) (Figure \ref{fig-pccc-csmf}). GPT-3.5 and GPT-4 had improvements ranging from 0.14-0.18 PCCC over InSilicoVA and InterVA-5, while GPT-4 slightly improved over GPT-3.5 by 0.05 PCCC. Population level performances were similar for all models (0.74-0.79 CSMF). Figure \ref{fig-pccc} shows the PCCC performance across three age groups (adult, child, and neonate). GPT-4 had the best individual performance for adult and neonatal records (0.64 and 0.58 PCCC), while GPT-3.5 had the best performance for child records (0.54 PCCC) with GPT-4 performing slightly worse (0.51 PCCC). InSilicoVA and InterVA-5 performed the worse for adult and child records ($\leq$0.5 PCCC), while GPT-3.5 performed the worse for neonatal records (0.42 PCCC). Across age ranges, all models followed a similar pattern in individual performance (Figure \ref{fig-pccc-age}). PCCC trended upwards for 1 month to 14 years ($\sim$0.1-0.75), and downwards for ages 15 to 69 years ($\sim$0.7-0.35). The highest and lowest performances were observed for ages 10-29 years ($\sim$) and 1-11 months respectively. Performances varied more across models for ages 0 days to 5 years, while less variation was seen between GPT-3.5 and GPT-4, as well as InSilicoVA and InterVA-5, from 5 to 69 years.

@@

At the population level, overall performances for all models were similar to physicians (0.74-0.79 CSMF), indicating potential for adequately estimating COD distributions for large populations. Although all models did not perform well for all records at the individual level (0.44-0.61 PCCC), several models performed well for certain CODs (0-0.99 PCCC). For most CODs, GPT-3.5/GPT-4 performed better than InSilicoVA/InterVA-5 (top PCCC for 15 of 17, 7 of 8, and 4 of 5 adult, child, and neonatal CODs respectively), while InSilicoVA performed better for particular CODs (road and transport injuries, tuberculosis, pneumonia, and neonatal infections with 0.84, 0.71, 0.78, and 0.86 PCCC respectively). For CODs with high performance (e.g. GPT-3.5/4 with 0.91-0.99 PCCC for maternal conditions, chronic respiratory disease, diabetes melitus, and cancers, InSilicoVA with 0.84 and 0.86 PCCC for road and transport injuries, and neonatal infections), the results suggest that GPT-3.5/4 and InSilicoVA may assign CODs that are very similar to physicians. Thus, it may be beneficial to evaluate performance at the COD level, and apply a combination of models that perform well in comparison to physicians for each COD. For example, different models perform well for various leading CODs as seen in Table \ref{tab-leadcod} \citep{worldhealthorganizationWorldHealthStatistics2024,worldhealthorganizationCausespecificMortality2000}.

@@

Across age ranges, all models followed a similar upward trend from 6 months to 14 years of age, and a downward trend from 15-69 years with GPT models having higher performance than InSilicoVA/InterVA-5 models, while more mixed trends were observed from 0 days to 5 months (recall Figure \ref{fig-pccc-age}). For adult age ranges, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults with some improvements after the age of 59. For child and neonatal age ranges, the performance improved drastically as the age increased after 5 months, suggesting less difficulty in COD assignment when children and neonates are more developed. As the models did not perform particularly well ($\geq$ 0.8 PCCC) for any specific five-year age range, it is not recommended to apply specific models that target cases by age. However, the patterns of increases and decreases of performance in relation to age provide valuable insight for comparison to expected physician diagnosis patterns in well-studied medical literature and knowledge. For example, it may be expected that physicians are more uncertain in diagnosing diseases that are prevalent in neonatal patients \citep{rasmussenComplexityPhysiciansUnderstanding2019,faisonWhenUnknownUnknowable2023}, which are present in our findings from Figure \ref{fig-pccc-age}.

@@

Recall that the GPT models in this study had the temperature parameter set to 0 for more reproducible and reliable results. A short experiment in Appendix \ref{app-reprod} revealed that GPT-3.5 assigns the same COD for the same record only more than 60\% of the time, based on repeated runs on a sample of 100 records. This suggests that GPT models do not always reliably assign identical CODs for the same case on multiple runs, which may pose issues in reproducibility and reliability. For example, GPT models may achieve correct COD assignments solely due to random chance, but are difficult to test with large numbers (e.g. 10,000) of reruns due to costs (e.g. costs increased 10 fold per record when rerun 10 times). In comparison, InterVA-5 and InSilicoVA are open source and free, allowing a large number of reruns without incurring additional fees. In addition, InterVA-5 and InSilicoVA assign CODs and provide probabilities for each alternative COD, which offers more reproducible and reliable COD assignments despite lower performance overall. Lastly, a major disadvantage in all models was that they were trained on historical data up to particular points in time, which may not utilize the most up-to-date data available (e.g. latest online articles, social media, or books for GPT models). Emergent diseases (e.g. COVID-19) and changes in distributions (e.g. outbreaks) may not be caught by these models depending on how often they are updated.

@@

GPT-3.5 and GPT-4 models required large computing infrastructure to train and run, which was not possible to run on local computers, or setup due to costs and ownership of the models. This poses issues with data privacy as sensitive data (e.g. identifying information) need to be sent to company servers, which can be collected by companies (e.g. OpenAI) and misused \citep{taoOpeningPandoraBox2023}. For example, in our study, GPT models use prompts, which contain the narrative data, to assign CODs, and the data in these prompts may be unknowingly collected and misused by companies (e.g. companies) or their users (e.g. malicious prompts) to identify participants or leak sensitive sensitive data \citep{khowajaChatGPTNeedsSPADE2024,wuUnveilingSecurityPrivacy2024}. In contrast, InterVA-5 and InSilicoVA can be run on local computers, which allows data to stay with the owner to protect data privacy, without reliance on external services.

@@

This section identifies limitations in this research in the context of GPT models. Section \ref{discuss-limits-icd10} identifies the omission of ICD-10 performance evaluations. Section \ref{discuss-limits-model} mentions the need for parameter tuning and evaluation of consistency and multiple COD assignments.

\subsubsection{ICD-10 Evaluation and Low Sample Sizes}\label{discuss-limits-icd10}

For the scope of this study, all models were evaluated for their performance in broad CGHR-10 COD categories as opposed to more specific ICD-10 codes. However, in practical cases, physicians assign more specific ICD-10 codes rather than broader COD categories. InterVA-5 and InSilicoVA assigned broader WHO VA codes, and were unable to assign ICD-10 codes, as the number of cases for specific ICD-10 codes are often low and inadequate for training statistical models. In relation, some broader CGHR-10 CODs were even removed for performance evaluation as \textless 10 cases were captured (e.g. congenital anomalies, suicide). Although GPT models were able to assign ICD-10 codes, lower performance may be expected as even physicians do not agree completely on ICD-10 codes, noted that broader categories (CMEA-10 codes in Additional file 2) were used to assign equivalency or agreement.

@@

GPT-3.5 and GPT-4 models used default parameters with the exception of setting the temperature to 0 for more consistent results. However, the temperature and other model settings may be adjusted to possibly improve performance for GPT models \citep{openaiOpenAIPlatformAPI2024}. This was not examined as sensitivity analyses on model parameters are costly across multiple reruns, noted in Section \ref{discuss-disadv-reprod}, which is required when testing various parameter settings. In addition, GPT models may possibly produce inconsistent results even with the temperature set to 0. Thus, it is important to also test the reliability and consistency of GPT outputs to avoid coincidental results due to randomness \citep{johnsonAssessingAccuracyReliability2023,jangConsistencyAnalysisChatGPT2023,krishnaEvaluationReliabilityRepeatability2024}. InterVA-5 and InSilicoVA were able to provide multiple COD assignments with probabilities for each COD. GPT models can be prompted to produce more than one COD assignment, but was not explored in this study as we only evaluated the most probable COD. This may be useful to evaluate the performance of multiple alternative COD assignments, which may provide additional diagnoses that have a higher chance of being similar to physician assignment, and better reflect causes leading to death \citep{solemanVerbalAutopsyCurrent2006}.

@@

Prompt engineering, the design of prompts to guide GPT models for better results \citep{wangPromptEngineeringHealthcare2024}, presents an important research opportunity that may improve performance of GPT models for COD assignment. An example exploration was conducted in Appendix \ref{app-misclass} on misclassified GPT-4 records for neonatal infections, which found potential issues with the categorization of CGHR-10 codes, order of information in narratives, and guidelines of COD assignments. An analysis of misclassified records with domain experts (e.g. physicians, specialists) may yield insights on adjusting prompts to assign more correct CODs, or apply more relevant broad COD categories for evaluation. In addition, subsequent prompts, data, and examples can be used to include correctional instructions and refine results, while additional information from the questionnaire and physician VA manuals can provide contextual information (e.g. retrieval augmented generation \citep{lewisRetrievalaugmentedGenerationKnowledgeintensive2020}) for further performance improvements \citep{meskoPromptEngineeringImportant2023}. Sensitivity analyses may be conducted to assess the effects on performance and consistency of results from modified prompts on a COD basis. GPT models may also be customized to specific domains or contexts, where objectives, behaviours, extra data, privacy, and evaluation tests can be adjusted to produce custom models that perform better in targeted domains or circumstances (e.g. custom models for particular CODs) \citep{almasreDevelopmentEvaluationCustom2024}.

\subsubsection{Guided and Monitored Household Surveys}\label{discuss-opp-gsurveys}

Recall that VAs involve surveyors that visit households to gather information about the deceased from their family, next-of-kin, friend, or community. Although standard questionnaires are used during this visit, there is significant information, containing latent patterns, from the narrative that is not always captured by the questionnaire \cite{kingQualityDiagnosticValue2016,jebleeAutomaticallyDeterminingCause2019}. These narratives often require a human connection between the surveyor and household members, where surveyor characteristics vary in social ability, cultural understanding, emotional capacity, and medical knowledge that affect the quality and bias of narratives \citep{solemanVerbalAutopsyCurrent2006,lohAddedValueOpen2021}. GPT models may help guide surveyors during VA interviews to probe households for better narrative information by generating and suggesting better questions, or providing questions that may have been missed by the surveyors. In addition, as models can assign CODs on-demand, there is potential for models to provide immediate COD estimates during the data collection process to monitor data quality on-demand (e.g. comparing estimated to expected COD distributions for known areas as quality checks).

@@

Our study lays the foundation for the integration of GPT, InterVA-5, and InSilicoVA models into VA systems to assist physicians in COD assignment. In dual-coded VA systems (described in Section \ref{methods-data}), two physicians are randomly assigned to each record and require second inspections of each other's assignment (reconciliation) and evaluation by a third more senior physician if their assignments do not agree. As mentioned in Section \ref{discuss-adv-hscav}, suggestion of alternative assignments from GPT and InSilicoVA models potentially reduces the disagreement between physicians, and ill-defined records, while allowing physicians to focus on more difficult records. Thus, model suggestions can be integrated into VA systems by presenting COD suggestions to physicians after their initial COD assignment, which allows them to consider alternative assignments and possibly revise their assignments based on the suggestions. At step 2 in Figure \ref{fig-discuss-phyassist}, GPT, InterVA-5, and InSilicoVA models can suggest COD assignments to consider, providing the option in step 2b to revise or proceed with their initial assignment. Our future work will be a first step in computer assisted verbal autopsy, assessing the effects of these model suggestions on improve VA data quality (e.g. increase in agreed records, reduction of ill-defined deaths). In preparation, we have integrated GPT-4, InterVA-5, and InSilicoVA model suggestions into our on-going HEAL-SL study after survey round 2 \citep{njalauniversityHealthySierraLeone2023} with goals of increasing physician agreement and reducing ill-defined COD assignments.

@@

This study evaluates the performance of GPT-3.5, GPT-4, InterVA-5, and InSilicoVA models compared to physicians for assigning CODs for 6939 VA records in Sierra Leone (2019-2022). At the population level, all models were similar (0.74-0.79 CSMF accuracy). At the individual level, GPT-4 had the best performance (0.61 PCCC), followed by GPT-3.5 (0.58 PCCC), and InSilicoVA/InterVA-5 (0.44 PCCC). Across CODs, GPT-4 had performed best for 10 of 17 adult, 4 of 8 child, and 3 of 5 neonatal CODs, with GPT-3.5 for 5 adult, 3 child, and one neonatal CODs, and InSilicoVA for 2 adult, one child, and one neonatal CODs. Model performance increased ($\sim$0.1-0.75 PCCC) as children and neonates developed (0 days to 14 years), and decreased ($\sim$0.7-0.35) as adults aged (15 to 69 years). Thus, GPT and InSilicoVA models were comparable to physicians for several CODs, but not across age ranges. As performance varied across CODs and age ranges, it is advantageous to combine several models to target CODs that each model performs well for, and to compare age-related performance patterns in relation to physicians. In addition, all models were able to scale to a large number of records and were available on-demand in comparison to physicians, enabling COD estimation and alternative diagnoses in low resource or physician scarce scenarios. As GPT models operate on natural language, they are able to adapt to more loosely defined data structures (e.g. assign in different COD coding standards, provide reasoning, and use contextual information when samples are low), making them behave more similarly to physician assignment. However, GPT models do not provide reliable CODs on repeated assignment, and were limited to past training data, with large computing infrastructure requirements, leading to reproducibility issues in COD assignments, difficulty adapting to new or changing CODs, and data privacy issues. Limitations of this study included difficulty comparing ICD-10 codes directly due to incompatible COD outputs from each model and low sample sizes, difficulty in conducting sensitivity analyses for GPT models due to costs, and omitting evaluation of multiple COD assignments due to study scope. We identified research opportunities in refining GPT models using prompt engineering and custom models for improving performance, guided household surveys to improve narrative quality, and future work in computer assisted VA, where GPT and other models will be used to assist physician COD assignment by offering multiple alternative assignments, with goals of increasing agreement on COD assignment and reducing ill-defined deaths. GPT-4, InterVA-5, and InSilicoVA has been integrated into future survey rounds of the HEAl-SL study from 2022 onwards, offering alternative COD assignments to assist physicians with second opinions. Future work in evaluating the effectiveness of computer assisted VA to reduce disagreements among physicians and ill-defined deaths will help support the advancement of more accurate and efficient VA systems across the world.

@@

\begin{figure}[H]
\centering
\includegraphics[width=.7\textwidth]{fig-methods.pdf}
\caption{Detailed case study methods.}\label{fig-methods}
\end{figure}
